---
output:
  pdf_document: default
  html_document: default
---

# Simple denoising examples

```{r}
library(tidyverse)
```

In this exercise:

1. First, we'll generate a time series of "measurements" where the process we are measuring behaves exactly like a sine wave.
   This will be our "ground truth" signal.
2. Next, we'll simulate different types of noise that could result from imperfect measurements.
3. Finally, we'll apply 2 basic denoising techniques to each of our noisy signals.

## Generating the ground truth

For this demonstration, we'll generate a signal that behaves like a sine wave.
Feel free to adjust any of the parameters to see how it influences our ability to denoise the signal.

```{r}
# Resolution gives the number of measurements in our signal.
resolution <- 1000
# x_range_min gives the minimum x value in our signal.
x_range_min <- 0
# x_range_max gives the maximum x value in our signal.
x_range_max <- (8 * pi)
# Generate our signal x,y:
x <- (seq(from=x_range_min, to=resolution) / resolution) * x_range_max
y <- sin(x)
# Create a data frame
data <- data.frame(x=x, y_true=y)
```

`data$y_true` gives the ground truth signal.

We can graph our ground truth:

```{r}
ggplot(data, aes(x = x, y = y_true)) +
  geom_point() +
  labs(
    x = "X",
    y = "True value",
    title = "Ground truth"
  )
```

## Adding some noise

First, let's simulate that we had a noisy sensor such that each measurement has a small amount of random noise.
To do so, we'll add a little bit of Gaussian noise to our ground truth signal.

```{r}
# noise_sd controls the level of noise applied to each measurement
noise_sd <- 0.1
data$y_noise <- data$y + rnorm(length(data$y), 0, 0.1)
```

We can graph this noisy signal:

```{r}
ggplot(data, aes(x = x, y = y_noise)) +
  geom_point() +
  labs(
    x = "X",
    y = "Noisy measurement value"
  )
```

Next, let's add another kind of noise that we might encounter with our sensor.
What if our sensor randomly fails, returning a nonsense value (in this case, let's say 99).

```{r}
prob_sensor_failure <- 0.02 # Sensor fails 2% of the time
failure_value <- 99
# "Simulate" when our sensor fails. Use rbinom to flip a biased coin for each object in our data.
data$sensor_failure <- rbinom(nrow(data), 1, prob_sensor_failure)
data$y_noise_faulty <- mapply(
  function(y, fail) {
    if (fail) {
      return(failure_value)
    } else {
      return (y)
    }
  },
  data$y_noise,
  data$sensor_failure
)
```

Graph our faulty sensor data.

```{r}
ggplot(data, aes(x = x, y = y_noise_faulty)) +
  geom_point() +
  labs(
    x = "X",
    y = "Faulty measurement value"
  )
```

We have our nice signal, and then we have those sensor failure measurements throwing everything off.

## Denoising using a rolling average

Let's try denoising using a rolling average.
This technique loops over each value in your sequence, setting that value to the average of its neighbors.
The number of neighbors is determined by the "window size".

```{r}
# This function calculates a rolling average over the sequence parameter.
# window_width defines how far the window extends to either side of the each index.
rolling_avg <- function(sequence, window_width=1) {
  avgs <- c()
  for (i in seq_along(sequence)) {
    window <- c(i)
    for (w_i in 1:window_width) {
      left <- i - w_i
      right <- i + w_i

      # Check that our window doesn't walk off the edge of our sequence.
      # If so, extend the other side of the window.
      if (left < 1) {
        left <- (i + window_width) + (abs(1 - left))
      }
      if (right > length(sequence)) {
        right <- (i - window_width) - ((right - length(sequence)) )
      }
      window <- c(window, left)
      window <- c(window, right)
    }
    avgs <- c(avgs, mean( sequence[window] ))
  }
  return(avgs)
}
```

Apply our rolling average function to `y_noise`

```{r}
window_w <- 10
data$y_noise_denoised <- rolling_avg(
  data$y_noise,
  window_width=window_w
)
ggplot(data, aes(x=x, y=y_noise_denoised)) +
  geom_point()
```

Much smoother! The rolling average works well when you want to smooth out a small magnitude of random noise.

Apply our rolling average function to `y_noise_faulty`:

```{r}
window_w <- 10
data$y_noise_faulty_denoised <- rolling_avg(
  data$y_noise_faulty,
  window_width=window_w
)
ggplot(data, aes(x=x, y=y_noise_faulty_denoised)) +
  geom_point()
```

Yikes! Our signal looks _worse_ than it did before we applied the rolling average denoising.

### Exercises

- Generate a small sequence of 10 random values.
  On paper, apply a rolling average filter to your sequence.
  You decide the window size.
  - What would happen if you increased your window size?
  - What about if you decreased your window size?
- Is the rolling average a good denoising approach for `y_noise`? What about `y_noise_faulty`?
- Did the rolling average improve how representative `y_noise_faulty` is of the ground truth signal?
  - Try calculating the [mean squared error](https://en.wikipedia.org/wiki/Mean_squared_error) for each of `y_noise` and `y_noise_denoised`. Does error go down?
  - What about for `y_noise_faulty` and `y_noise_faulty_denoised`?
- What happens when you make the rolling average window smaller? Bigger?
- What happens when you add _more_ noise to the signal?
- Identify any lines of code that you don't understand. Use the documentation to figure out what those lines of code are doing.

## Denoising using a median filter

A median filter is similar to applying a rolling average,
except instead of replacing each value in our sequence with the average of its neighbors, we replace it with the median of its neighbors.

```{r}
# This function calculates a rolling average over the sequence parameter.
# window_width defines how far the window extends to either side of the each index.
median_filter <- function(sequence, window_width=1) {
  avgs <- c()
  for (i in seq_along(sequence)) {
    window <- c(i)
    for (w_i in 1:window_width) {
      left <- i - w_i
      right <- i + w_i

      # Check that our window doesn't walk off the edge of our sequence.
      # If so, extend the other side of the window.
      if (left < 1) {
        left <- (i + window_width) + (abs(1 - left))
      }
      if (right > length(sequence)) {
        right <- (i - window_width) - ((right - length(sequence)) )
      }
      window <- c(window, left)
      window <- c(window, right)
    }
    avgs <- c(avgs, median( sequence[window] ))
  }
  return(avgs)
}
```

Apply our rolling average function to `y_noise`:

```{r}
window_w <- 10
data$y_noise_denoised <- median_filter(
  data$y_noise,
  window_width=window_w
)
ggplot(data, aes(x=x, y=y_noise_denoised)) +
  geom_point()
```

Apply our rolling average function to `y_noise_faulty`:

```{r}
window_w <- 10
data$y_noise_faulty_denoised <- median_filter(
  data$y_noise_faulty,
  window_width=window_w
)
ggplot(data, aes(x=x, y=y_noise_faulty_denoised)) +
  geom_point()
```

### Exercises

- Do you notice any differences in the results of applying a median filter versus applying a rolling average?
- Generate a small sequence of 10 random values.
  On paper, apply a median filter to your sequence. You decide the window size.
  - What happens if you increase your window size?
  - What about if you decrease your window size?
- Is the median filter a good denoising approach for `y_noise`? What about `y_noise_faulty`?
  - Try calculating the [mean squared error](https://en.wikipedia.org/wiki/Mean_squared_error) for each of `y_noise` and `y_noise_denoised`. Does error go down?
  - What about for `y_noise_faulty` and `y_noise_faulty_denoised`?
- Did the median filter improve how representative `y_noise_faulty` is of the ground truth signal?
- What happens when you make the median filter window smaller? Bigger?
- What happens when you add _more_ noise to the signal?
- Identify any lines of code that you don't understand. Use the documentation to figure out what those lines of code are doing.

## Further exploration

As we talked about in class, the appropriate-ness of different denoising techniques will depend on your data and your particular application domain.
The R community has implemented many denoising techniques.
For example, [the `signal` package](https://www.rdocumentation.org/packages/signal/versions/0.7-7) includes many signal denoising functions.

