---
output:
  pdf_document: default
  html_document: default
---

# Dealing with missing values

In this lab activity, we will use R to deal with missing values in a dataset.
There are many methods of dealing with missing values in a dataset.
Here, we'll cover how to apply a few simple/basic approaches in R:

- Feature elimination (remove columns with missing values)
- Object elimination (remove rows with missing values)
- Mean imputation
- Median imputation

This activity is meant to server as a "how to" guide and does not discuss _when_ different methods of handling missing values should be applied.
Whether or not a particular approach is appropriate will depend on your particular dataset and domain.

## Dependencies and setup

We'll need to load the tidyverse packages for this lab activity:

```{r}
library(tidyverse)
```

## Testing for missing values

In R, missing values are often represented by `NA` (or some other value particular to your data; e.g., -1, none, etc).
You can identify missing values using `is.na()` (run `?is.na` to see documentation).

In this demo, we'll play with the `airquality` dataset, which comes loaded automatically in R.

```{r}
head(airquality)
```

There are `NA` values in the `airquality` dataset.
To get a binary dataframe that indicates which positions have `NA` values, you can:

```{r}
airquality_na <- is.na(airquality)
head(airquality_na)
```

## Dropping columns with missing values (feature elimination)

First, let's get a list of columns that contain missing values.

```{r}
columns_with_na <- c()
for (col in colnames(airquality)) {
  if (NA %in% airquality[,col]) {
    columns_with_na <- c(columns_with_na, col)
  }
}
print(columns_with_na)
```

You can verify that against the `airquality_na` matrix we created earlier.

To drop columns, we can use the `select` function in dplyr.

```{r}
dropped_cols_data <- airquality %>%
  select(
    !all_of(columns_with_na)
  )
head(dropped_cols_data)
```

## Dropping rows with missing values (object elimination)

If you know exactly what columns have missing values in them, you can use `filter` to filter out rows where those columns have NA.

```{r}
dropped_rows_data <- airquality %>%
  filter(
    !(is.na(Ozone) | is.na(Solar.R))
  )
head(dropped_rows_data)
```

A little bit more generic, we can use the matrix we generated earlier, `airquality_na`.

```{r}
# First, let's figure out all of the row ids that we want to drop

drop_rows <- c()
for (ri in 1:nrow(airquality_na)) {
  if (TRUE %in% as.vector(airquality_na[ri,])) {
    drop_rows <- c(drop_rows, ri)
  }
}

# We can use filter to drop all of the row ids in drop_rows
dropped_rows_data2 <- airquality %>%
  filter(
    !(row_number() %in% drop_rows)
  )
head(dropped_rows_data2)
```

## Mean imputation

A mean imputation replaces missing values for a particular feature with the mean of other values for that feature.
We can use `dplyr` functions to pretty easily perform a mean imputation on a particular column:

```{r}
airquality %>%
  mutate(
    Ozone=ifelse(is.na(Ozone), mean(Ozone, na.rm=TRUE), Ozone)
  )
```

In the code above, we mutate the Ozone column. If the value is na, we replace it with the mean of the Ozone column, and otherwise, we leave it as is.

## Median imputation

A median imputation replaces missing values for a particular feature with the median of other (non-missing) values for that feature.
We can do the exact same thing we did for a mean imputation, but swap mean out for median:

```{r}
airquality %>%
  mutate(
    Ozone=ifelse(is.na(Ozone), median(Ozone, na.rm=TRUE), Ozone)
  )
```

## Exercises

- Describe a procedure for replacing an object's missing value for a particular feature with the value from its "nearest neighbor"?
  - What does it mean for one object (i.e., row) to be "near" another object? How could you quantify how close two objects are to one another?
  - How would you implement this approach in R? (either from scratch or using an existing function in a library you install/load)
- Assume that you have a dataset where each object has two features: a continuous feature (where some of the values are missing) and a categorical feature.
  In this dataset, objects of the same category are more similar to each other than to objects in different categories.
  You want to apply a mean imputation to this dataset to replace the missing values in the continuous feature.
  How could you use each object's category to improve your mean imputation?