---
output:
  pdf_document: default
  html_document: default
---

# Building a decision tree in R

In this lab activity, we will use the `rpart` R package to build a decision tree classifier.

## Dependencies and setup

We'll use the `rpart`, `rpart.plot`, and `caret` packages in this activity.
If you do not have these packages installed, you will need to do so (use the `install.packages` command).

```{r, warning=FALSE}
library(rpart) # Used to build decision tree model
library(rpart.plot) # Used to visualize trees
library(caret) # Used to build a fancy confusion matrix
```

I encourage you to look up the documentation for these packages.

We'll also use the following packages:
```{r}
library(tidyverse)
```

Next, we'll load the dataset that contains data for the 100 US senators as of summer 2020.
**Depending on your working directory, you may need to adjust the path to the data file.**

```{r}
senator_data <- read_csv("lecture-material/week-06/senate-data.csv")
senator_data$pol <- as.factor(senator_data$pol)
senator_data$law <- as.factor(senator_data$law)
senator_data$mil <- as.factor(senator_data$mil)
senator_data$soc <- as.factor(senator_data$soc)
senator_data$edu <- as.factor(senator_data$edu)
senator_data$med <- as.factor(senator_data$med)
senator_data$bus <- as.factor(senator_data$bus)
senator_data$agr <- as.factor(senator_data$agr)
senator_data$fin <- as.factor(senator_data$fin)
senator_data$approval <- as.factor(senator_data$approval)
head(senator_data)
```

These data were taken from a website of senator rankings.
Each row in these data describes a senator with the following attributes:

- `first` - First name
- `last` - Last name
- `pol` - Political party (democrat, republican, or independent)
- `age` - Age of the senator
- `law` - Binary attribute describing whether senator had a previous career in law enforcement
- `mil` - Binary attribute describing whether senator had a previous career in the military
- `soc` - Binary attribute describing whether senator had a previous career in social services
- `edu` - Binary attribute describing whether senator had a previous career in education
- `med` - Binary attribute describing whether senator had a previous career in medicine
- `bus` - Binary attribute describing whether senator had a previous career in business
- `agr` - Binary attribute describing whether senator had a previous career in aggriculture
- `fin` - Binary attribute describing whether senator had a previous career in finance
- `years` - Number of years in office
- `approval` - Approval class. class=0 means voters don't approve of them, and class=1 means that voters do approve of them.

## Building a decision tree

We can use the `rpart` function to build a decison tree:

```{r}
model <- rpart(
  formula = approval ~ pol + age + law + mil + soc + edu + med + bus + agr + fin + years,
  data = senator_data,
  parms = list(split="information")
)
model
```

If you've never seen the formula syntax before, I recommend reading up on it a bit: <https://r4ds.had.co.nz/model-basics.html?q=formula#formulas-and-model-families>
In the formula used to build our decision tree model, `approval` is our dependent variable and all of the predictor attributes to the right of the `~` are our independent variables.

We can use the `summary` function for a little bit more information about our model:

```{r}
summary(model)
```

## Visualizing our decision tree

We'll use functionality from the `rpart.plot` package to visualize the decision tree we built:

```{r}
# If you want to save the visualization, uncomment the next few lines
pdf("decision_tree.pdf")
rpart.plot(model)
dev.off()
# (this isn't using ggplot, so we can't use ggsave)

# Redraw plot to make it show up in our Rmd doc
rpart.plot(model)
```

## Evaluating the accuracy of our decision tree

Our goal with this model was primarily descriptive.
That is, we wanted to learn more about the properties of our senators dataset instead of building a model to predict the approval rating of unknown senators.

As such, we did not create an explicit training and testing set.
In essence, the entire senator dataset is our "training set".
So, while we don't have a testing set to evaluate our model, we can evaluate our model's accuracy on the training set.

```{r}
ground_truth <- senator_data$approval
training_predictions <- predict(
  model,
  data=senator_data,
  type="class"
)
confusion_matrix <- table(ground_truth, training_predictions)
confusion_matrix
```

We can also use the `confusionMatrix` function from the `caret` package to build a fancier confusion matrix for us.

```{r}
confusionMatrix(confusion_matrix)
```

## Using our decision tree to make a prediction

Let's pretend that a mysterious new senator shows up to join the US senate one day (ignoring any and all laws that would prevent this from happening in reality).
```{r}
# Make an empty dataframe w/correct attributes to hold new senator.
new_senator <- senator_data[0,0:13]
# Bind new mystery senator to empty dataframe
new_senator <- rbind(
  new_senator,
  list(
    fist="Mystery",
    last="Senator",
    pol="I",
    age=42,
    law=0,
    mil=0,
    soc=0,
    edu=0,
    med=0,
    bus=0,
    agr=0,
    fin=0,
    years=0
  )
)
# Turn new senator columsn into factors where appropriate
new_senator$pol <- as.factor(new_senator$pol)
new_senator$law <- as.factor(new_senator$law)
new_senator$mil <- as.factor(new_senator$mil)
new_senator$soc <- as.factor(new_senator$soc)
new_senator$edu <- as.factor(new_senator$edu)
new_senator$med <- as.factor(new_senator$med)
new_senator$bus <- as.factor(new_senator$bus)
new_senator$agr <- as.factor(new_senator$agr)
new_senator$fin <- as.factor(new_senator$fin)
# I know that was all a little annoying just to predict a hand-crafted example.
# Usually the things you want to predict are in a file, so it's less annoying to
# load them in from file...
```

```{r}
result <- predict(
  model,
  newdata=new_senator,
  type="class"
)
result[[1]]
```

It looks like we would predict that Mystery Senator would not have voter approval (according to our model).

## Exercises

- Identify any lines of code that you do not understand. Use the documentation to figure out what is going on.
- Describe the decision tree. What is the most informative attribute for determining whether or not a senator has voter approval?
- How would you describe the performance of our model? What is its training error?
- Try adjusting the attributes for the Mystery Senator we created. What attributes can you change to change our model's prediction?
- Identify a dataset that might be interesting to build a decision tree model for.
  - Load that dataset into R
  - Partition the data into a training and testing set (i.e., split the rows)
  - Use rpart to build a decision tree using the training set.
  - What is the accuracy of your model on the training set?
  - What is the accuracy of your model on the testing set?
