---
title: "R Data Example"
author: "Alexander Lalejini"
date: "2022-09-01"
output: html_document
---

# Setup

```{r}
library(tidyverse)
library(viridis)  # Contains some nice color palettes
library(hexbin)   # Used to make the hexbin plot
```

# Load our data

```{r}
# Note that I'm using the read_csv function from the tidyverse readr package
# instead of the base r read.csv function. You can use either.
netflix_data <- read_csv("netflix_titles.csv")
```

Columns in our dataset:

```{r}
colnames(netflix_data)
```

First, let's make the `type`, `rating`, and `country` columns into factors.

```{r}
# Make type column a factor
netflix_data$type <- as.factor(netflix_data$type)

# Make ratings into a factor, throw out movies/shows with nonsense for a rating
netflix_data <- netflix_data %>% filter( !(rating %in% c("66 min", "74 min", "84 min")) )
netflix_data$rating <- as.factor(netflix_data$rating)

# Country into a factor
netflix_data$country <- as.factor(netflix_data$country)
```

Next, we split the data into two data sets, one for movies and one for tv shows.

```{r}
movie_data <- netflix_data %>% filter( type == "Movie" )
tv_show_data <- netflix_data %>% filter( type == "TV Show" )
```

Pull out the movie length (in minutes) from the duration column.

```{r}
movie_data <- movie_data %>%
  separate(
    col=duration,
    into=c("length_min", NA),
    sep=" ",
    remove=FALSE
  )
# make the new length_min column a numeric column
movie_data$length_min <- as.numeric(movie_data$length_min)
```

Pull out the tv show number of seasons from the duration column

```{r}
tv_show_data <- tv_show_data %>%
  separate(
    col=duration,
    into=c("num_seasons", NA),
    sep=" ",
    remove=FALSE
  )
tv_show_data$num_seasons <- as.numeric(tv_show_data$num_seasons)
```

We can use `group_by` and `summarise` to make a summary table that gives the number of movies on netflix per year.
Each observation will be a year with the number of available movies that year.

```{r}
# Use group_by to group movies by release_year, and then summarize will run some summary functions on them.
# These summary functions (e.g., n() and mean()) need to be compatible with summarise.
movie_release_year_table <- movie_data %>%
  group_by(release_year) %>%
  summarise(
    num_movies=n(),
    avg_length_min=mean(length_min, na.rm=TRUE)
  )
movie_release_year_table
```

Show the years with the greatest representation on netflix

```{r}
movie_release_year_table %>% arrange(desc(num_movies))
```

# Creating graphics with R

Histogram of number of movies on netflix by year

```{r}
# Graph adatped from R Graph Gallery
# https://r-graph-gallery.com/220-basic-ggplot2-histogram.html

my_plot <- ggplot(movie_data, aes(x=release_year)) +
  geom_histogram(binwidth=1) +
  labs(x="Release year", y="Number of movies")
my_plot

ggsave(plot=my_plot, filename="my_plot.pdf")
```

2-D density plot showing the relationship between release year and length of movie in minutes.

```{r}
# Graph adapted from the R Graph Gallery
# https://r-graph-gallery.com/2d-density-plot-with-ggplot2.html#hex

# Bin size control + color palette
ggplot(movie_data, aes(x=release_year, y=length_min) ) +
  geom_hex(bins = 70) +
  scale_fill_continuous(name="Number of movies", type = "viridis") +
  labs(y="Movie length (minutes)", x="Release year")
  theme_bw()
```


