---
output:
  pdf_document: default
  html_document: default
---

# Interactive k-means clustering example

```{r,echo=FALSE}
demo_url <- "https://lalejini.com/k-means-web-demo/web/kmeans_clustering.html"
```

Below is an interactive k-means clustering demo. [You may also access the demo here](https://lalejini.com/k-means-web-demo/web/kmeans_clustering.html).

In the demo, the black square is your data "canvas".
Click the "drop points" button to populate the canvas with random points.
You can also place points on the canvas manually by clicking to drop a point.
The clear button will clear all of the currently drawn points.
Configure `k` to set the number of clusters.

Press "cluster mode" to cluster.
You can run kmeans step-by-step by repeatedly pressing the "step" button.
You can tell the demo to just run kmeans continuously with the "run" button.
You can still add points to the canvas by clicking in cluster mode.
Press "data mode" to go back to configuring.

```{r,echo=FALSE}
knitr::include_url(demo_url, height="800px")
```